---
title: "Research Seminar - Logistic Regression"
author: "Johannes, Kai"
output: 
  github_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preamble

This README file serves as an outline of today's topics and contains all required materials or links to the materials. This file and all materials can be found on github. Today's aim is

  - to explain the structure of the **data** for which a logistic regression can be performed, 
  - to explain the special properties of the **logistic regression** and thereby introduce the concept of generalized regression, and
  - to **perform and interpret** a logistic regression in `R`.
  - At the end, examples from **current research projects** are shown and discussed.
  
The martial contains texts and examples from the textbook Fahrmeir et al. (2013) as well as parts and ideas from the corresponding lecture `GLM` by Prof. Thomas Kneib, Chair of Statistics. Furthermore, parts of the lectures `Statistical Data Analysis with R` and `Advanced Statistical Programming` of the forest faculty are included.

We are using the following libraries:

```{r, message=FALSE}
library(tidyverse)  # Data science
library(ggplot2)    # Visualization

# GLM itself is part of the stats package, which is loaded at start up
library(ggeffects)  # Visual interpretation of statistical models
library(emmeans)    # Testing linear hypotheses
```

We ask you to prepare some tasks beforehand to save some time in the seminar. Those are tagged as 'Beforehand task'.

__Beforehand task 0:__ Getting ready

- Connect to the repository and clone it to your laptop
- Get R and R Studio running
- Install the required libraries

Modelling survival probabilities, also known as survival analysis, is one of the most important and most suitable applications for which logistic regression is used and which we will also use as an example today. Another famous field of application is the questionnaire analysis, which will be sketched in the PdD examples at the end. The special property of survival data is that the response variable can only have binary (or more generally categorical) outcomes. Individuals will survive until one specific time point. Idea of survival modelling is to estimate this time point and possibly additionally inference the reasons survival (or non-survival respectively). This 0/1 property is the reason, why the values of the response that are to be estimated have different outcomes than the observations of the response variable. In logistic regression, not the outcome of an observation (e.g. dead or alive) is estimated, but the probability that an observation shows an outcome (e.g. suvival). Logistic regression is a possibility to fit a survival probability curve over a continuous variable of interest, e.g. the tree age. Logistic regression is thus an alternative to the well-known Weibull survival curves. The logit curve resulting from the regression has similar properties as the cumulative Weibull distribution. Also the interpretation is comparable.

# Data: `ForestHealth`

Consider the example data set `forest health` from Fahrmeir et al. (2013). The data set consists of 16 variables with 1796 observations on forest health to identify potential factors influencing the health status of trees and therefore the vital status of the forest. In addition to covariates characterizing a tree and its stand, the exact locations of the trees are known. The interest is on detecting temporal and spatial trends while accounting for further covariate effects in a flexible manner. The data in our example come from a specific project in the forest of Rothenbuch (Spessart), which has been carried out by Axel Göttlein (Technical University, Munich) since 1982. Five tree species are part of this survey: beech, oak, spruce,
larch, and pine. Here we will restrict ourselves to beech trees. Every year, the condition of beech trees is categorized by three ordinal categories \[0 \% - 12.5\), \[12.5 \% - 50 \%\), and \[50 \% - 75 \%\). The forth category >75 \% is not captured in the data. The category 0 % signifies that the beech tree is healthy, the category 100 % implies that the tree is dead. See p. 9 of Fahrmeir et al. (2013) for more details.

```{r}
load("Data/ForestHealth.rda")
ForestHealth %>% ggplot(aes(y = defoliation, x = age, col = stand)) + geom_point()
```

__Beforehand Task 1:__ Getting familiar with the data and understand the nature of binary responses

- Load `ForestHealth`
- Calculate a binary response variable that contains healthy (defoliation \[0 \% - 12.5\)) and unhealthy (\[12.5 \% - 75 \%\)) trees
- Perform relevant descriptive statistics
- Which variables might have an impact on the tree health?
- Fit an ordinary linear model `lm` to estimate the health status over the age
- Add one another variable that seems to influence the tree health
  -   Are these two models suitable to estimate the health status? Give some pro and contra arguments and underpin your arguments using common numbers or diagrams of (linear) statistical inference.

```{r}
ForestHealth <- ForestHealth %>% mutate(diseased = ifelse(defoliation %in% "[0, 12.5)", 0, 1))
ForestHealth %>% ggplot(aes(y = diseased, x = age)) + geom_point() +
  geom_smooth(method = "lm")
lm_simple <- ForestHealth %>% lm(diseased ~ age, data = .)
summary(lm_simple)
lm_multiple <- ForestHealth %>% select(age, elevation, soil, ph, fertilized, diseased) %>%  lm(diseased ~ ., data = .)
summary(lm_multiple)
ggeffects::ggeffect(lm_multiple, terms = "age") %>% plot()
```


# Orinary LM vs. GLM

Motivating example for GLMs: Why is the ordinary linear models usually not suited to model binary responses (and under which conditions could a linear model might be sufficient)?

__Beforehand Task 2:__ What is the differnce between linear models and general linear models? What is the G in GLM?

- Read Lane (2002) and prepare for the following discussion
    - Why do generalized linear model belong to the family of linear model even though their curvature is not linear?
    - Which properties distinguish a *generalized linear model* from a *linear* model? In particular: What is a link function?
    - Which properties distinguish a *generalized linear model* from a *non-linear* model?


# GLM

@Johannes, ich habe schon Mal was zu GLM vorbereitet, siehe misc. Das ist aber zu viel. Mich würde deine Meinung interessieren, ob es sich lohnt davon wsa zu recyclen?

## Logistic regression

Instead of modelling $y_i$ directly, we define a regression model that estimates the possibility of a discrete status 1 (e.g. dead)

$$
\pi_i=\text{E}(y_i)=P(y_i=1),
$$

where $\pi_i$ shall vary continuously between 0 and 1 even tough $y_i$ is discrete. To achieve $\pi_i \in[0, 1]$, we apply a transformation (in the sense of Lane (2002)) that transforms ("links") the linear regression formula to the deserved properties. The linear regression formula (so to speak the linear core of the GLM)

$$
\eta_i=\beta_0+\beta_1 x_{i1}+ \dots + \beta_k x_{ik}
$$

is called linear predictor. Instead of using a quite simple link function as Lane (2002) did (the logaritm), binomial regression requires a more sophisticated transformation

$$
\pi_i=h(\eta_i).
$$

Along with some less familiar transformations, the logit transformation leads to the desired properties, which is why the binomial is sometimes referred to as logit regression. The term logistic regression comes from the logistic curvature that follows from the logit transformation. The logit response function $h$ relies on the logistic cumulative distribution function

$$
\pi=h(\eta)=\frac{\text{exp}(\eta)}{1+\text{exp}(\eta)}.
$$

The respective inverse, also called link function $h$ reads as

$$
g(\pi)=\text{log} \left( \frac{\pi}{1 - \pi}\ \right).
$$

The response function follows a Bernoulli distribution. However, grouping observation with same outcomes makes the solving process more efficient and then leads to a Binomial distribution (Fahrmeir et al., 2023, p. 270 - 277), which is why in practiced programming, the Binomial or Quasibinomial distributions are usually taken. You should stick to these standard combination of distributions of link functions, however, you can freely decide for any of the distributions of the so called exponential family if want.

@Johannes, würdest du Logits per Hand rechnen lassen und dann in lm? Um in der Logik von dem Lane Paper zu beleiben?

__Task 3:__ Modeling a disease probability model 

- Calculate Logits by hand ???
- Create a univariate GLM over tree age
- Create a multiple GLM over age and self-chosen promising covariates, save separate to the simple GLM

```{r}
ForestHealth %>% ggplot(aes(y = diseased, x = age)) + geom_point() +
  geom_smooth(method = "glm", method.args = list(family = binomial(link = "logit")))

glm_simple <- ForestHealth %>% glm(diseased ~ age, data = ., family = binomial(link = "logit"))
summary(glm_simple)
glm_multiple <- ForestHealth %>% select(age, elevation, soil, ph, fertilized, diseased) %>% 
  glm(diseased ~ ., data = .,  family = binomial(link = "logit"))
summary(glm_multiple)
ggeffects::ggeffect(glm_multiple, terms = c("age")) %>% plot()
ggeffects::ggeffect(glm_multiple, terms = c("elevation")) %>% plot()
ggeffects::ggeffect(glm_multiple, terms = c("soil")) %>% plot()
```

## Model diagnostics

Von uns vorgestelltes Summary gemeinsam erklären
Kreuzvalidierung?
Overdispersion (quasibinomial, use more covariates)

## Model interpretaion

Von uns vorgestelltes Summary gemeinsam erklären
Odds multiplicative -> hard to interprete


# Outlook: Multinomial regression

Regression models that extend models for binary responses $y_i \in \{0, 1\}$ to the case of categorical responses $y_i=\{1, \dots, c+1\}$. In the example data `ForestHealth`, we have $y_i\in\{\text{no defoliation}, \text{weak defoliation}, \text{severe defoliation} \}$.

## Unordered catagories

In case of nominally scaled responses (no ordering), the logit model is extended with

## Ordered categories

Text und Code verfassen und nur vorstellen, keine Übungsaufgabe

    - Forest health multinomial
    - Forest health ordered
    
## PhD examples

- Henning
- Valeska

# Literature and Material

This file as well as all other materials are uploaded on GitHub. You should all have ssh access rights, such that you can use your version control feature of RStudio to clone the GitHub repository: 

https://github.com/Forest-Economics-Goettingen/Research_Seminar_GLM
git@github.com:Forest-Economics-Goettingen/Research_Seminar_GLM.git


## Primary literature

Lane (2002): Generalized linear models in soil science. European Journal of Soil Science, 53, 241-251. https://doi.org/10.1046/j.1365-2389.2002.00440.x

## Online resources

Johannes Signer, Kai Husmann (2024): Kursmaterial: Einführung in die Datenanalyse mit R. https://github.com/Forest-Economics-Goettingen/KursskriptRBsc

## Books

Ludwig Fahrmeir, Thomas Kneib, Stefan Lang, Brian Marx (2013): Regression : Models, Methods and Applications. Berlin, Heidelberg. Springer. https://link.springer.com/book/10.1007/978-3-642-34333-9