---
title: "Research Seminar - Logistic Regression"
author: "Johannes, Kai"
output: 
  github_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preamble

This README file serves as an outline of today's topics and contains all required materials or links to the materials. Today's aim is

  - to explain the nature of the data for which a logistic regression can be performed, 
  - to explain the special properties of the logistic regression and thereby introduce the concept of generalized geression, and
  - to perform and interpret a logistic regression in `R`.
  - At the end, examples from current research projects are shown and discussed.
  
The martial contains texts and examples from the textbook Fahrmeir et al. (2013) and also parts and didactic concepts are borrowed from the corresponding lecture `GLM` by Prof. Thomas Kneib, Chair of Statistics. Also parts of the lectures `Statistical Data Analysis with R` and `Advanced Statistical Programming` are included.

We are using the following libraries:



Modelling survival probabilities, also known as survival analysis, is one of the most important applications for which logistic regression is used and which we will use as an example today. Another famous field of application is the questionnaire analysis, which will be sketched in the PdD examples at the end. The special property of survival data is that the response variable can only have binary (or more generally categorical) outcomes. This property is the reason, why the values of the response that are to be estimated have different outcomes than the observations of the response variable. In logistic regression, not the outcome of an observation (e.g. dead or alive) is estmimated, but the probability that an observation shows an outcome. Logistic regression is a possibility to fit a survival probability curve over a continuous variable of interest, e.g. the tree age. Logistic regression is therefore an alternative to the well-known Weibull survival curves. The logit curve resulting from the regression has similar properties as the cumulative Weibull distribution. Also the interpretation is comparable.

# Data: `Forest Health`

Consider the example data set `forest health` from Fahrmeir et al. (2013). The data set consists of 16 variables with 1796 observations on forest health to identify potential factors influencing the health status of trees and therefore the vital status of the forest. In addition to covariates characterizing a tree and its stand, the exact locations of the trees are known. The interest is on detecting temporal and spatial trends while accounting for further covariate effects in a flexible manner. The data in our example come from a specific project in the forest of Rothenbuch (Spessart), which has been carried out by Axel Göttlein (Technical University, Munich) since 1982. Five tree species are part of this survey: beech, oak, spruce,
larch, and pine. Here we will restrict ourselves to beech trees. Every year, the condition of
beech trees is categorized by the response variable “defoliation” (`defol`) into nine ordinal
categories 0 %, 12.5 %, 25 %, 37.5 %, 50 %, 62.5 %, 75 %, 87.5 %, and 100 %. Whereas the
category 0 % signifies that the beech tree is healthy, the category 100 % implies that the tree
is dead. See p. 9 of Fahrmeir et al. (2013) for more details.

```{r}
load("Data/ForestHealth.rda")
library(ggplot2)
library(tidyverse)
```



# Literature and Material
Johannes Signer, Kai Husmann (2024): Kursmaterial: Einführung in die Datenanalyse mit R. https://github.com/Forest-Economics-Goettingen/KursskriptRBsc

Ludwig Fahrmeir, Thomas Kneib, Stefan Lang, Brian Marx (2013): Regression : Models, Methods and Applications. Berlin, Heidelberg. Springer. https://link.springer.com/book/10.1007/978-3-642-34333-9